---
title: "Overview of biotidyr"
author: "Phil Chapman"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette: 
    toc: true
vignette: >
  %\VignetteIndexEntry{Overview of the biotidyr package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE, warning=FALSE}
library(dplyr)
library(PharmacoGx)
library(biotidyr)
library(ggplot2)
# note - how to export as pdf rmarkdown::render('vignettes/Overview.Rmd', 'pdf_document', output_dir='~/Desktop')
```


## Introduction

Exemplify functions for generating tidy data from PharmacoSet and MultiAssayExperient objects.

## Quick example

A PharmacoSet object contains molecular profiling and drug response data for collections of cell lines.  The molecular profiling data is stored in an ExpressionSet object

```{r}
data('CCLEsmall', package='PharmacoGx')
CCLEsmall
eset <- CCLEsmall@molecularProfiles$rna
class(eset)
```

The `gather.ExpressionSet` function converts an ExpressionSet into a tidy data frame:

```{r}
df <- gather.ExpressionSet(eset, sample_ids=c('143B', '23132-87'), gene_ids=c('BRAF', 'EGFR'), 
                           sample_col = "cellid", gene_col = "Symbol")
df
as.data.frame(df)
```

Not specifying gene\_ids or sample\_ids returns all genes or samples

```{r}
gather.ExpressionSet(eset, sample_ids=c('143B', '23132-87'), 
                           sample_col = "cellid", gene_col = "Symbol")
```

Note that we can customise which identifier we want to use by specifying the sample\_col and gene\_col parameters.

## gather.PharmacoSet

The `gather.PharmacoSet` function combined data from multiple sources and reports in a tidy format with a single data point per row:

```{r}
gather.PharmacoSet(CCLEsmall, sample_ids=c('143B', '23132-87'), gene_ids=c('BRAF', 'EGFR'),
  data_types=c('rna', 'mutation'), gene_col=c('Symbol', 'Symbol'))
```

## make\_genetic\_\vs\_genetic_df.PharmacoSet

The `make_genetic_vs_genetic_df.PharmacoSet` function generates a data frame with one pair of genetic feature data per row.

```{r}
gvg_df <- make_genetic_vs_genetic_df.PharmacoSet(CCLEsmall, sample_ids=cellNames(CCLEsmall), gene1='RBM5', gene2='RBM5',
data_type1='rna', data_type2='rnaseq', gene_col1 = "Symbol", gene_col2 = "gene_name")
gvg_df

```

This is useful for creating plots of one gene vs another in ggplot2:

```{r}
ggplot(gvg_df, aes(x=feature_value1, y=feature_value2)) + 
    geom_point() + 
    theme_bw()

```

However, you can also supply multiple genes to this function:
```{r}
genes <- c('RBM5', 'NQO1', "STPG1", "NIPAL3","LAS1L","ENPP4","SEMA3F","CFTR")
gvg_df2 <- make_genetic_vs_genetic_df.PharmacoSet(CCLEsmall, sample_ids=cellNames(CCLEsmall), gene1=genes,  gene2=genes, data_type1='rna', data_type2='rnaseq', gene_col1 = "Symbol", gene_col2 = "gene_name")
nrow(gvg_df2)

```

This data format is then in a convenient form to manage models in a data frame using the dplyr and broom packages - see [Managing Many Models by Hadley Wickham](https://www.youtube.com/watch?v=rz3_FDVt9eg).  In the example below we are exploring how well RNAseq and Affymetrix data correlate across a number of genes.

```{r}
mod_df <- gvg_df2 %>%
    dplyr::filter(gene1==gene2) %>%
    dplyr::group_by(feature_name1, feature_name2) %>%
    tidyr::nest() %>%
    dplyr::mutate(mod=purrr::map(data, function(x) lm(feature_value1 ~ feature_value2, data=x)),
                  res=purrr::map(mod, broom::glance))
mod_df
mod_df %>% dplyr::select(-data,-mod) %>% tidyr::unnest()

```


