---
title: "Overview of tidyMultiAssay"
author: "Phil Chapman"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette: 
    toc: true
vignette: >
  %\VignetteIndexEntry{Overview of the tidyMultiAssay package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE, warning=FALSE}
library(dplyr)
library(PharmacoGx)
library(tidyMultiAssay)
library(ggplot2)
# note - how to export as pdf rmarkdown::render('vignettes/Overview.Rmd', 'pdf_document', output_dir='~/Desktop')
```


## Introduction

Exemplify functions for generating tidy data from PharmacoSet and MultiAssayExperient objects.

## Quick start

A PharmacoSet object contains molecular profiling and drug response data for collections of cell lines.  The molecular profiling data is stored in an ExpressionSet object

```{r}
data('CCLEsmall', package='PharmacoGx')
CCLEsmall
eset <- CCLEsmall@molecularProfiles$rna
class(eset)
```

The `gather.ExpressionSet` function converts an ExpressionSet into a tidy data frame:

```{r}
df <- gather.ExpressionSet(eset, sample_ids=c('143B', '23132-87'), gene_ids=c('BRAF', 'EGFR'), 
                           sample_col = "cellid", gene_col = "Symbol")
df
as.data.frame(df)
```

Not specifying gene\_ids or sample\_ids returns all genes or samples

```{r}
gather.ExpressionSet(eset, sample_ids=c('143B', '23132-87'), 
                           sample_col = "cellid", gene_col = "Symbol")
```

Note that we can customise which identifier we want to use by specifying the sample\_col and gene\_col parameters.

The `gather.PharmacoSet` function combined data from multiple sources and reports in a tidy format with a single data point per row:

```{r}
gather.PharmacoSet(CCLEsmall, sample_ids=c('143B', '23132-87'), gene_ids=c('BRAF', 'EGFR'),
  data_types=c('rna', 'mutation'), gene_col=c('Symbol', 'Symbol'))
```

The `make_genetic_vs_genetic_df.PharmacoSet` function generates a data frame with one pair of genetic feature data per row.

```{r}
gvg_df <- make_genetic_vs_genetic_df.PharmacoSet(CCLEsmall, sample_ids=cellNames(CCLEsmall), gene1='RBM5', gene2='RBM5',
data_type1='rna', data_type2='rnaseq', gene_col1 = "Symbol", gene_col2 = "gene_name")
gvg_df

```

This is useful for creating plots of one gene vs another in ggplot2:

```{r}
ggplot(gvg_df, aes(x=feature_value1, y=feature_value2)) + 
    geom_point() + 
    theme_bw()

```

However, you can also supply multiple genes to this function:
```{r}
genes <- c('RBM5', 'NQO1', "STPG1", "NIPAL3","LAS1L","ENPP4","SEMA3F","CFTR")
gvg_df2 <- make_genetic_vs_genetic_df.PharmacoSet(CCLEsmall, sample_ids=cellNames(CCLEsmall), gene1=genes,  gene2=genes, data_type1='rna', data_type2='rnaseq', gene_col1 = "Symbol", gene_col2 = "gene_name")
nrow(gvg_df2)

```

This data format is then in a convenient form to manage models in a data frame using the dplyr and broom packages - see [Managing Many Models by Hadley Wickham](https://www.youtube.com/watch?v=rz3_FDVt9eg).  In the example below we are exploring how well RNAseq and Affymetrix data correlate across a number of genes.

```{r}
mod_df <- gvg_df2 %>%
    dplyr::filter(gene1==gene2) %>%
    dplyr::group_by(feature_name1, feature_name2) %>%
    tidyr::nest() %>%
    dplyr::mutate(mod=purrr::map(data, function(x) lm(feature_value1 ~ feature_value2, data=x)),
                  res=purrr::map(mod, broom::glance))
mod_df
mod_df %>% dplyr::select(-data,-mod) %>% tidyr::unnest()

```

## Data formats

There are four data frame formats used in the tidyMultiAssay package:

- tall\_df: Common currency minimal data frame which can contain any data type (see `data(example_tall_df)`)
- resp\_df: Used to store user specified response data from a cell panel screen or siRNA screen (see `data(dietlein_data)`)
- gvg\_df: Genetic vs Genetic data frame used to model or plot one set of genetic features against another set (see `data(example_gvg_df)`)
- rvg\_df: Response vs Genetic data frame used to model or plot a set of response data against a set of genetic features (see `data(example_rvg_df)`)

Since the same data frame format with the same column names can be used for limitless combinations of genetic and response data, it is easy to create reusable analysis and visualisation functions.  

```{r}
#tall_df example
data(example_tall_df)
dplyr::tbl_df(example_tall_df)

#resp_df example
data(dietlein_data)
dplyr::tbl_df(dietlein_data)

#gvg_df example
data(example_gvg_df)
dplyr::tbl_df(example_gvg_df)

#rvg_df example
data(example_rvg_df)
dplyr::tbl_df(example_rvg_df)

```

## Combining data of different types

Converting data in a PharmacoSet or MultiAssayExperiment object into a standardised data frame format makes combining data from different sources very simple since one can just be appended to another to create another tall_df formatted data frame, or one can be joined to another to create the response vs genetic or genetic vs genetic data frames.  By following the principles of tidy data, feature data is stored in rows rather than columns to make the data frames as generic and reusable as possible.  

For example, response and assay data can be gathered seperately from a PharmacoSet object using the `gather_response.PharmacoSet` and `gather_response.PharmacoSet` functions :
```{r}
resp_data <- gather_response.PharmacoSet(CCLEsmall, sample_ids = NULL, resp_ids = c('lapatinib', 'Erlotinib', 'PLX4720'))
assay_data <- gather_assay.PharmacoSet(CCLEsmall, sample_ids=NULL, gene_ids = c('BRAF', 'ERBB2', 'EGFR', 'NQO1'), 'rna', gene_col='Symbol')
```

Or in a single data frame via a command using the `gather.PharmacoSet` function:
```{r}
combined_data <- gather.PharmacoSet(CCLEsmall, sample_ids=NULL, data_types='rna', 
                                    gene_col='Symbol', gene_ids = c('BRAF', 'ERBB2', 'EGFR', 'NQO1'),
                                    resp_col = "ic50_published", resp_ids = c('lapatinib', 'Erlotinib', 'PLX4720'))
```

The gather.PharmacoSet function can also be used to extract data from multiple assay types:
```{r}
combined_data2 <- gather.PharmacoSet(CCLEsmall, sample_ids=NULL, data_types=c('rna', 'rnaseq'), 
                                    gene_col=c('Symbol', 'gene_name'), gene_ids = c('RBM5', 'NQO1', "STPG1", "NIPAL3","LAS1L","ENPP4","SEMA3F","CFTR"))
```


These data frames can then be combined using the generic `make_genetic_vs_genetic_df.data.frame` and `make_response_vs_genetic_df.data.frame` functions:

```{r}
gvg1 <- make_genetic_vs_genetic_df.data.frame(assay_data)
rvg1 <- make_response_vs_genetic_df.data.frame(assay_data, resp_data)
```

Note that by default a number of arguments for each function are set to NULL.  More control over which genes and samples are included can be gained by setting these parameters, for example:
```{r}
gvg2 <- make_genetic_vs_genetic_df.data.frame(combined_data2, gene1=c('RBM5', "STPG1"), gene2=c('RBM5', "STPG1"))

```

Alternatively, the resultant data frame can be filtered to make downstream analyses more useful:
```{r fig.width=6, fig.height=6}
gvg3 <- gvg2 %>%
    dplyr::filter(gene1 == gene2, feature_type1=='rna', feature_type2=='rnaseq')

ggplot(gvg2, aes(feature_value1, feature_value2)) +
    geom_point() + geom_smooth(method='lm') +
    facet_grid(feature_name1~feature_name2)

ggplot(gvg3, aes(feature_value1, feature_value2)) +
    geom_point() + geom_smooth(method='lm') +
    facet_grid(feature_type1+gene1~feature_type2)

```
It is also possible to do the entire operation in one function call using the `make_genetic_vs_genetic_df.PharmacoSet` function.

## Combining data from different data sources
Data from different PharmacoSet or MultiAssayExperiment objects can be combined by first converting it to the tall\_df data frame format and then using `make_genetic_vs_genetic_df.data.frame` or `make_response_vs_genetic_df.data.frame` functions.

```{r}
ccle_assay_data <- gather_assay.PharmacoSet(CCLEsmall, sample_ids=NULL,                                             gene_ids = c('RBM5', "STPG1", 'CFTR', 'MTMR7'), data_type='rna', gene_col='Symbol')

data('GDSCsmall', package='PharmacoGx')
gdsc_assay_data <- gather_assay.PharmacoSet(GDSCsmall, sample_ids=NULL,                                             gene_ids = c('RBM5', "STPG1", 'CFTR', 'MTMR7'), data_type='rna', gene_col='Symbol')


gvg4 <- make_genetic_vs_genetic_df.data.frame(df=ccle_assay_data, df2=gdsc_assay_data)
gvg5 <- gvg4 %>% dplyr::filter(gene1==gene2)
ggplot(gvg5, aes(feature_value1, feature_value2)) +
    geom_point() + geom_smooth(method='lm') + xlab('CCLE') + ylab('GDSC') +
    facet_grid(~gene1) + theme_bw()

```

## Converting sample ids

When combining data from different sources, there is no guarantee that the sample identifies used are the same.  In particular this is an issue when using cell line data from different experiments, although the PharmacoSet object provided in the PharmacoGx package do an excellent job of unifying the cell line id's.

If sample identifiers do need to be unified, the `convert_ids` function can be used.  This takes any data frame with a column named sample\_id (for example a tall\_df) and another data frame with at least two columns, one of which represents the sample id pre-conversion, and the other representing the sample id post-conversion.  An example data frame is provided with this package that provides conversion information for a variety of identifier schemas for cancer cell lines, but user generated ones can also be used:

```{r}
data("CancerCellLineIDs")
CancerCellLineIDs
convert_ids(df=example_tall_df, id_data=dplyr::filter(CancerCellLineIDs, id_type=='CCLE'), from_col='alt_id', to_col='unified_id')
```

The function can also be used to append additional columns to the resultant data frame by setting the `other_cols` parameter.

## Using custom data

All that is required is to put the custom data frame into either the tall\_df or resp\_df data frame format and then use either the `check_df_format` or `get_df_format` function to ensure that the format is correct:

```{r}
test_df <- dplyr::data_frame(sample_id=c('A375_SKIN', 'DMS114_LUNG', 'NCIH1155_LUNG'),
                             assayed_id=rep('BRAF', 3),
                             data_type=rep('qPCR', 3),
                             original=as.character(c(1,2,3)),
                             value=c(1,2,3))
get_df_format(test_df)
check_df_format(test_df, 'tall_df')
```

Once the data is in the right format, it can be combined with data that originates from PharmacoSet or MultiAssayExperiment objects as described in the previous section on combining data from different sources.  In the example the ids also need to be converted:

```{r}
data("dietlein_data")
gdat1 <- gather.PharmacoSet(CCLEsmall, gene_ids = c('BRAF', 'EGFR', 'TP53'), gene_col = c('Symbol', 'Symbol'), data_types = c("rna", "mutation"))
gdat1 <- gdat1 %>% convert_ids(id_data=dplyr::filter(CancerCellLineIDs, id_type=='CCLE'), from_col='alt_id', to_col='unified_id')

#now create response_vs_genetic
dietlein_rvg <- make_response_vs_genetic_df.data.frame(gdat1, dietlein_data)
dietlein_rvg

```







